What is an Ancestral Recombination Graph?

### 2022-02-18 paper.tex

Formally, we can define a GARG as a set of edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section).

@@

\section*{Ancestry resolution and simplification}
% This is a rough first pass. Will clarify terminology later.
The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (or ``marginal'') trees.

@@

% This is weak. Should we give a better description of the algorithm?
% Not clear yet, and don't want to spend too much time on it if it's
% unhelpful.
Given a set of edges $(u, v, I)$ where node $u$ descends from $v$
over the set of genomic intervals $I$, and a set of sample nodes $S$
it is straightforward to construct the local tree at position $x$.
Starting from the sample nodes in $S$, traverse upwards through the
graph, such that when we are at a node $u$ we follow the edge
$(u, v, I)$ where $x \in I$. This is essentially the same procedure as
described by Griffiths~\citep{griffiths1997ancestral}.

Recovering the tree at a single point along the genome, although
a fundamental operation, is not the only operation that we may wish
to perform on an ARG. We can make the operation of efficiently
constructing \emph{all} trees along the genome much more efficient,
and gain insight into the structure of the ARG,
if we perform an additional step and \emph{resolve} the ancestral
material in the ARG with respect to the samples.

As we saw in the section X, the key to efficiently accessing the
results of the ARG stochastic process is to reason about
the flow of ancestral material through the graph using Hudson's
algorithm. The same basic process allows us to determine the
exact intervals of ancestral material that are flowing through
each edge in the graph. Concretely, given a set of edges
$E = \{(u, v, I)\}$ where $u$ is the child, $v$ is the parent
and $I$ is the set of genomic intervals passed directly from
$v$ to $u$, the resolution process with respect to a set of
samples $S$ produces a set of edges $E' = \{(u, v, I)\}$
where $I$ is the set of genomic intervals over which there is a
% weak and not correct - how do we say this properly?
path from at least one element of $S$ to $u$,
and $u$ is not an MRCA.

@@

@@

>>>>

Formally, we can define a GARG as a set of edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section). Deriving the local tree at a point $x$
follows largely the same pattern as for an EARG but is somewhat
simpler because coordinate information is directly associated with
edges. Starting with the sample nodes $S$ we traverse
upwards through the graph from child to parent. At each node $u$, we find an
edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$ (if it
exists) and move to the next node $p$.

@@

\section*{Ancestry resolution and simplification}
% This is a rough first pass. Will clarify terminology later.
The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitrarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (or ``marginal'') trees.

@@

Given a GARG $E$ and the set of samples $S$ we can \emph{resolve} its
ancestry by noting that the resolved interval $Ir$ on an edge $(c,p,I)$
is given by [Rough notes]

\begin{itemize}
\item The ``ancestry interval'' $A_u$ of a node $u$ is the union of
all $R_{u, v}$ for all the children $v$ of $u$. (this isn't quite right)
\item The resolved interval is then the intersection of $A_u$ and the
original inheritance interval $I_v$ on the edge.
\item It would be good of if differentiate between the inheritance intervals
 from the GARG and the ancestry intervals of the resolved GARG.
\end{itemize}

@@

% Not sure where this text will end up, but we want it somewhere
% towards the end.

@@

% TODO not clear this belongs in the discussion, but we'll definitely
% want to make some of these points, even if they are being repeated
% from earlier section.
It is difficult to imagine a practical calculation performed on an
ARG that does not involve either the sequential generation of trees
along the genome or the resolved ancestry segments.
Knowing the resolved segments in an ancestry resolved GARG allows
us to perform many calculations without explicitly generating
the trees, and lets us work with the ancestry on a genome-by-genome
basis. We give three examples here. Firstly, generating mutations
on edges can be performed very quickly on an ancestry resolved
GARG, since we only need to examine $O(|E|)$ edges rather than
$O(nt)$ tree branches (assuming $n$ samples and $t$ marginal trees)
~\citep{baumdicker2021efficient}. If the ancestry were not resolved,
we would generate a large number of mutations on areas of the edge that
are non-ancestral. Secondly, computing the likelihood of a given
ARG realisation under the coalescent with recombination model
requires that we essentially recapitulate the state of Hudson's
algorithm (see XXX section) at each node in
the graph~\citep{baumdicker2021efficient}. Having resolved ancestry
substantially simplifies this process. Thirdly, computing
[FIXME describe what we did in tsdate]~\citep{wohns2021unified}.

Sequentially generating the marginal trees along the genome
is also fundamental, and is necessary whenever we need to
perform calculations that are contingent on more than just the
isolated properties of an edge. \cite{kelleher2016efficient}
showed how all trees can be sequentially generated in
constant time per tree transition in a fully simplified GARG.
Furthermore, we can easily reason about how tree topologies
change (and stay the same), leading to efficient algorithms
for computing population genetic
statistics~\citep{kelleher2016efficient,kelleher2018efficient},
implementing the Li and Stephens
model~\citep{kelleher2019inferring,wohns2021unified}
and likely many more.
